---
title: "Sydney Trains"
output:
  html_document:
    df_print: paged
---

# Overview

Analyse and visualise trips on the Sydney Trains Network. 

The data is from the realtime GTFS data at  `https://api.transport.nsw.gov.au/v1/gtfs/schedule/sydneytrains` that is assumed to sit in the `data` folder as unzipped files, and renamed to `.csv`.

# Setup

```{r, message=FALSE}
rm(list=ls())

library(tidyverse)
library(forcats)
library(stringr)

src_folder <- "data"
theme_set(theme_minimal())
```

# Import

```{r}
read_routes <- function () { read_csv(file.path(src_folder, "routes.csv")) }
read_stops <- function () { read_csv(file.path(src_folder, "stops.csv")) }
read_shapes <- function () { read_csv(file.path(src_folder, "shapes.csv")) }
read_trips <- function () { read_csv(file.path(src_folder, "trips.csv")) }
read_stop_times <- function () { read_csv(file.path(src_folder, "stop_times.csv")) }
```

# Transform

```{r}
transform_routes <- function (df) {
  df %>%
    filter(
      agency_id == "SydneyTrains",
      !is.na(route_short_name)) %>%
    select(
      route_id, 
      route_short_name,
      route_desc,
      route_long_name,
      route_color) %>%
    mutate(route_color = paste0("#", route_color))
}

transform_trips <- function (df) {
  df %>%
    select(
      route_id,
      trip_id,
      trip_headsign,
      direction_id,
      shape_id)
}

transform_stop_times <- function (df) {
  df %>%
    select(
      trip_id,
      stop_id,
      stop_sequence) %>%
    mutate(stop_id = as.character(stop_id))
}

transform_stops <- function (df) {
  df %>%
    mutate(
      stop_title = stop_name %>% 
        str_remove(" Station Platform \\d+") %>% 
        str_remove(" Station")) %>%
    select(
      stop_id,
      stop_code,
      stop_title,
      stop_lat, 
      stop_lon)
}
```


# Load

Read the data and do repeated joins to create an aggregate data frame.

```{r message=FALSE}
routes <- 
  read_routes() %>% 
  transform_routes

# Identify just the Sydney Trains routes
sydney_routes <- 
  routes %>%
  pull(route_id) %>%
  unique()

# Match only on route_id, and filter out non-Sydney Trains routes
trips <-
  read_trips() %>%
  transform_trips() %>%
  filter(route_id %in% sydney_routes)

# Match only on stop_id
stops <- 
  read_stops() %>% 
  transform_stops

# Match only on trip_id
stop_seq <-
  read_stop_times() %>%
  transform_stop_times %>%
  left_join(stops, by = "stop_id") %>%
  right_join(trips, by = "trip_id") %>%
  select(-stop_code, -stop_lat, -stop_lon)

all_stops <- 
  stop_seq %>%
  pull(stop_title) %>%
  sort() %>%
  unique()
```

View the structure of all major data frames.

```{r}
glimpse(routes)
glimpse(trips)
glimpse(stops)
glimpse(stop_seq)
```


# Analyse

Some questions to ask:

* How many distinct trips per line?
* What are all the stops on a given trip? 
* How many stops on each line?
* What is the overall distance of each line?
* What stations are common to different routes?


## Distinct trips

Get all the routes and number of trips

```{r}
trips_per_route <- 
  trips %>%
  inner_join(routes, by = "route_id") %>%
  group_by(route_long_name, route_color) %>%
  tally(name = "total_trips")
```


Extract colours for each long name.

```{r}
route_colours <- 
  routes %>%
  arrange(route_long_name) %>%
  select(route_desc, route_color) %>%
  distinct()

#%>%
#  pull(route_color)
```


How many distinct trips per route?

```{r}
trips_per_route %>%
  ggplot(
    aes(x = reorder(route_long_name, total_trips), 
        y = total_trips)) +
  geom_col(
    aes(fill = as_factor(route_color))) +
  geom_text(
    aes(label = total_trips), 
    size = 3, 
    hjust = -0.2, 
    colour = "#999999") +
  coord_flip() +
  scale_y_continuous(
    limits = c(0, 4000), 
    breaks = seq(0, 4000, 500)) +
  scale_fill_manual(
    values = route_colours$route_color) +
  labs(
    title = "Total number of distinct trips on each route", 
    subtitle = "Sydney Trains",
    x = NULL, 
    y = "Count") +
  theme(legend.position = "none")

```

## Stops per trip

Return the list of stations for a given trip ID, or a given route ID.

```{r}
stops_by_trip_id <- function (id) {
  stop_seq %>%
    filter(
      trip_id == id) %>%
    select(
      trip_id,
      stop_id, 
      stop_sequence) %>%
    distinct() %>%
    inner_join(stops, by = "stop_id") %>%
    pull(stop_title)
}

stops_by_route_id <- function (id) {
  trips %>%
    filter(route_id == id) %>%
    head(n=1) %>%
    pull(trip_id) %>%
    stops_by_trip_id()
}
```


Examples

```{r warning=FALSE}
stops_by_trip_id("1--A.1259.126.128.B.8.59842229")
```

```{r}
stops_by_route_id("IWL_1c")
```


Count the number of stops

```{r warning=FALSE}
routes %>%
  select(
    route_id, 
    route_long_name) %>%
  inner_join(trips, by = "route_id") %>%
  group_by(route_id) %>%
  mutate(
    number_of_stops = stops_by_route_id(route_id) %>% length()) %>%
  select(
    route_id,
    route_long_name,
    number_of_stops) %>%
  distinct() %>%
  arrange(desc(number_of_stops)) %>%
  head(10)
```

## Stop locations

Show the location of all stops

```{r}
stop_seq %>%
  select(stop_id) %>%
  distinct() %>%
  left_join(
    stops, by = "stop_id") %>%
  select(
    stop_title, stop_lon, stop_lat) %>%
  ggplot(
    aes(x = stop_lon, y = stop_lat)) +
  geom_point(
    colour =  "#3333ff",
    size = 0.5) +
  labs(
    title = "All stop locations",
    x = "Longitude", y = "Latitude")
```


## Graph of stops and edges

```{r}
#TODO
```


## Stops vs routes

### What routes does each stop serve? 

```{r}
stop_routes <-
  stop_seq %>%
  select(stop_title, trip_id) %>%
  left_join(trips, by = "trip_id") %>%
  left_join(routes, by = "route_id") %>%
  select(stop_title, route_desc, route_color) %>%
  distinct() %>%
  arrange(stop_title)
```

```{r}
glimpse(stop_routes)
```


### Top number of routes per stop

```{r}
stop_routes %>%
  group_by(stop_title) %>%
  summarise(routes = n()) %>%
  ggplot(aes(routes)) +
  stat_count(fill = "#33cc99", width = 0.7)

#  arrange(desc(routes)) %>%
#  head(8)
```



### Stops per route

```{r}
stops_per_route <-
  stop_routes %>%
  group_by(route_desc, route_color) %>%
  summarise(stops = n()) %>%
  arrange(desc(stops))

stops_per_route
```

```{r}
```


```{r}
stops_per_route %>%
  ggplot(aes(reorder(route_desc, stops), stops)) +
  geom_col(
    aes(fill = route_desc), 
    width = 0.6) +
  geom_text(
    aes(label = stops), 
    size = 3, 
    hjust = -0.5) +
  scale_fill_manual(
    values = deframe(route_colours)) +
  coord_flip() +
  theme(legend.position = "none")
```


```{r}
stop_colours <-  
  stop_routes %>%
  select(-route_desc) %>%
  group_by(stop_title) %>%
  summarise(
    #colours = as.list(route_color) %>% as_vector("character")
    colours = str_c(route_color, collapse = ",") %>% str_split(","))
    #colours = str_c(route_color, collapse = "|"))

stop_colours %>% head(10)
```

### Export

```{r}
stop_routes %>%
  write_csv("data/stop_routes.csv")
```

```{r}
library(jsonlite)
stop_colours %>%
  write_json("data/stop_colours.json")
```

## Endpoints

All trip starts and ends

```{r}
trip_endpoints <- function (df) {
  starts <- 
    df %>%
    group_by(route_id, trip_id) %>%
    filter(stop_sequence == 1) %>%
    select(route_id, trip_id, start_stop = stop_name)
  
  df %>%
    group_by(route_id, trip_id) %>%
    filter(stop_sequence == max(stop_sequence)) %>%
    select(route_id, trip_id, end_stop = stop_name) %>%
    inner_join(starts)
}
```


Get all the unique trips across the network as endpoints, regardless of times
```{r}
trip_ends <- 
  stops %>% 
  trip_endpoints %>%
  ungroup(trip_id) %>%
  select(-trip_id) %>%
  unique() %>%
  arrange(route_id)
```

# Visualise

Show a graph of the trips between endpoints

```{r}
library(igraph)

plot_endpoints <- function (df = trip_ends) {
  all_ends <- c(trip_ends$start_stop, trip_ends$end_stop) %>% unique
  gr <- 
    trip_ends %>%
    select(-route_id) %>%
    graph_from_data_frame(directed = TRUE, vertices = all_ends)

  deg <- gr %>% degree(mode = "all")
  layout <- 
    layout_with_lgl(gr) %>% 
    norm_coords(ymin=-1, ymax=1, xmin=-1, xmax=1)
  
  gr %>% 
    plot(vertex.size = log(deg) * 2,
         # vertex.label.cex = 0.3,
         vertex.color = "#6666ff",
         vertex.label = NA,
         edge.color = "#CC6666",
         edge.arrow.size = 0,
         rescale = FALSE,
         layout =  layout * 1.0)
}
  
```





```{r}
trip_ends %>% plot_endpoints
```

